### 2019-03-22 IEEE_TC_submission.d/abstract.tex
ssr

\vspace*{-0.05in}
\begin{center}
\textbf{Abstract}  
\end{center}

\smallskip

A class of Boolean functions, 
called \textbf{nested canalyzing functions} (NCFs),
has been used to model certain biological phenomena.
Researchers have studied various properties of these functions
such as sensitivity and stability.
The purpose of this note is to bring out some relationships between NCFs, symmetric 
Boolean functions and a generalization of symmetric Boolean functions,
which we call $r$-symmetric functions (where $r$ is the symmetry level).
Using a normalized representation for NCFs, we develop a 
characterization of when two variables of an NCF are symmetric.
Using this characterization, we show 
that the symmetry level of an NCF $f$
can be easily computed given a standard representation of $f$.
(In contrast, we prove that even approximating the symmetry level of
a general Boolean function to within any factor $\geq 1$ is \cnp-hard.) 
We also show that for any NCF $f$ with $n$ variables, the notion of
strong asymmetry considered in the literature is equivalent to
the property that $f$ is $n$-symmetric. 
We use this result to derive a closed form expression for the
number of $n$-variable Boolean functions 
that are NCFs and strongly asymmetric.
We also identify all the Boolean functions that are NCFs 
and symmetric.

>>>>

A class of Boolean functions, 
called \textbf{nested canalyzing functions} (NCFs),
has been used to model certain biological phenomena.
Researchers have studied various properties of these functions
such as sensitivity and stability.
The purpose of this note is to bring out some relationships between NCFs, symmetric 
Boolean functions and a generalization of symmetric Boolean functions,
which we call $r$-symmetric functions (where $r$ is the symmetry level).
Using a normalized representation for NCFs, we develop a 
characterization of when two variables of an NCF are symmetric.
Using this characterization, we show 
that the symmetry level of an NCF $f$
can be easily computed given a standard representation of $f$.
(In contrast, we prove that even approximating the symmetry level of
a general Boolean function to within any factor $\geq 1$ is \cnp-hard.) 
We also present an efficient algorithm for testing whether 
a given $r$-symmetric function is an NCF.
Further, we show that for any NCF $f$ with $n$ variables, the notion of
strong asymmetry considered in the literature is equivalent to
the property that $f$ is $n$-symmetric. 
We use this result to derive a closed form expression for the
number of $n$-variable Boolean functions 
that are NCFs and strongly asymmetric.
We also identify all the Boolean functions that are NCFs 
and symmetric.

### 2019-03-22 IEEE_TC_submission.d/intro.tex
ssr

Our focus is on the relationships between 
NCFs and symmetric Boolean functions.
Using a normalized representation for NCFs 
(defined in Section~\ref{sec:prelim}), we develop a
characterization of when two variables of an NCF are symmetric.
Using this characterization, we show
that the symmetry level of an NCF $f$
can be easily computed given a normalized representation of $f$.
(In contrast, we show that 
one cannot even efficiently approximate the symmetry level of
a general Boolean function to within any factor $\geq 1$, 
unless \textbf{P} = \cnp.)
We also show that for any NCF $f$ with $n$ variables, the notion of
strong asymmetry considered in the literature is equivalent to
the property that $f$ is $n$-symmetric.
We use this result to derive a closed form expression for the
number of $n$-variable Boolean functions
that are NCFs and strongly asymmetric.
In addition, we identify all the Boolean functions that are 
canalyzing and symmetric as well as those that 
are NCFs and symmetric.

>>>>

Our focus is on the relationships between 
NCFs and symmetric Boolean functions.
Using a normalized representation for NCFs 
(defined in Section~\ref{sec:prelim}), we develop a
characterization of when two variables of an NCF are symmetric.
Using this characterization, we show
that the symmetry level of an NCF $f$
can be easily computed given a normalized representation of $f$.
(In contrast, we show that 
one cannot even efficiently approximate the symmetry level of
a general Boolean function to within any factor $\geq 1$, 
unless \textbf{P} = \cnp.)
We also present an efficient algorithm to test whether a given
$r$-symmetric function $f$ is an NCF. 
Further, we show that for any NCF $f$ with $n$ variables, the notion of
strong asymmetry considered in the literature is equivalent to
the property that $f$ is $n$-symmetric.
We use this result to derive a closed form expression for the
number of $n$-variable Boolean functions
that are NCFs and strongly asymmetric.
In addition, we identify all the Boolean functions that are 
canalyzing and symmetric as well as those that 
are NCFs and symmetric.

### 2019-03-22 IEEE_TC_submission.d/symmetry_properties.tex
ssr

\subsection{Overview}
\label{sse:res_overview}
Our first result (Theorem~\ref{thm:ncf_symmetric_variables})
provides a characterization of pairs of variables of an NCF
that are part of the same symmetry group. 
This characterization allows us to give a simple closed form
expression (Theorem~\ref{thm:ncf_r_symmetric}) 
for the symmetry level of an NCF specified using
its default-normalized representation.
%% We also list some corollaries that follow from the
%% characterization.
Next, we show that for any NCF $f$ with $n$ variables,
the two statements ``$f$ is strongly asymmetric" ~and~
``the symmetry level of $f$ is $n$"~ are equivalent 
(Theorem~\ref{thm:ncf_strong_asymmetry}).
We use this result to derive a closed form expression 
for the number of $n$-variable NCFs that are 
also strongly asymmetric (Theorem~\ref{thm:count_strongly_asymmetric}).
Our final result (Proposition~\ref{pro:ncf_symmetric})
identifies all the functions that are symmetric and
canalyzing as well as those that are symmetric and NCFs.

@@

Theorem~\ref{thm:ncf_r_symmetric} shows that given an NCF $f$,
the problem of finding the smallest integer $r$ such that
$f$ is $r$-symmetric can be solved efficiently.
We now show that the converse problem, that is, determining whether a given
$r$-symmetric Boolean function is an NCF,
can also be solved efficiently.
When the answer is ``yes", our algorithm also constructs a
default-normalized representation of $f$.

Let $f(x_1, x_2, \ldots, x_n)$ be an $r$-symmetric Boolean function of $n$
Boolean variables.
We assume that the symmetry groups $g_i$, $1 \leq i \leq r$,
of $f$ are given.
Let $m_i$ denote the number of variables
in the group containing variable $x_i$, $1 \leq i \leq m$.
Thus, in the group containing $x_i$, the number
of variables which can take on the value 1 varies from 0
to $m_i$, $1 \leq i \leq n$.
We also assume that $f$ is
given as a table $T$ with $\mu = \prod_{i=1}^r (m_i+1)$ rows.
Each row of $T$ specifies an $r$-tuple $(c_1, c_2, \ldots, c_r)$,
where $c_i$ is the number of variables of group $g_i$ which
have the value 1, along with the \{0,1\} value of $f$ for that $r$-tuple.
As will be seen, our algorithm for determining whether $f$ is
an NCF runs in $O(r^2\mu)$ time.
The algorithm relies on the following observation.

\begin{observation}\label{obs:rsym_canalyzing}
Suppose $f(x_1, x_2, \ldots, x_n)$ is an $r$-symmetric Boolean function.
Consider any variable $x_i$ and suppose
the number of variables in the group
containing $x_i$ is $m_i$, $1 \leq i \leq r$.
\begin{enumerate}
\item The function $f$ is canalyzing in variable $x_i$
with canalyzing value 1 iff all the table entries where the number
of 1's in the group containing $x_i$ is nonzero have the same value
of $f$. (This value of $f$ is the canalyzed value when $x_i = 1$.)
\item The function $f$ is canalyzing in variable $x_i$ with
canalyzing value 0 iff all the table entries where the number of
1's in the group containing $x_i$ is less than $m_i$ have the same
value of $f$. (This value of $f$ is the canalyzed value when $x_i = 0$.)
\end{enumerate}
\end{observation}
We now specify an iterative algorithm for determining whether 
$f$ is an NCF, and if so, constructing a simplified 
NCF representation (defined in Section~\ref{sec:intro}) for $f$.
In iteration $j$, the algorithm uses a set $X_j$ of remaining variables,
table $T_j$ with $\mu_j$ rows, and the number of groups $g_j$.
Initially (i.e., $j = 1$), $X_1$ = $\{x_1, \ldots, x_n\}$,
$T_1$ is the given table $T$ for $f$, 
$\mu_1 = \mu$ (the number of rows of $T$) and $g_j = r$ (the number of
symmetry groups of $f$) and the NCF representation for $f$ is empty.
The algorithm executes as long as one or more symmetry groups remain.
In each iteration, the algorithm carries out the following steps.

\smallskip
\noindent
I. Use Observation~\ref{obs:rsym_canalyzing} to determine whether there is
a variable $x_i \in X_j$ is canalyzing.

@@

\smallskip
\noindent
The correctness of the algorithm is a direct consequence of
Observation~\ref{obs:rsym_canalyzing}. If the algorithm is successful,
it produces a simplified NCF representation of $f$; this can then be
efficiently converted into the default-normalized representation as
discussed in Section~\ref{sse:ncf_layer}.

To estimate the running time, we note that in iteration $j$, we can
determine in $O(g_j\,\mu_j)$ = $O(r\,\mu)$ time using 
Observation~\ref{obs:rsym_canalyzing}
whether there is a variable that is canalyzing.
(This can be done by considering one arbitrary variable 
from each group.) 
Once a variable is identified, the other steps in that iteration
can be completed in $O(\mu_j)$ = $O(\mu)$ time.
Thus, the time for each iteration is $O(r\,\mu)$.
Since there are at most $r$ iterations, the running time of the
algorithm is $O(r^2\,\mu)$.
The following theorem summarizes this result.

\begin{theorem}\label{thm:rsym_canalyzing}
Given an $r$-symmetric function $f$, the problem of testing whether
$f$ is an NCF can be solved efficiently. \QED
\end{theorem}

>>>>

\subsection{Overview}
\label{sse:res_overview}
Our first result (Theorem~\ref{thm:ncf_symmetric_variables})
provides a characterization of pairs of variables of an NCF
that are part of the same symmetry group. 
This characterization allows us to give a simple closed form
expression (Theorem~\ref{thm:ncf_r_symmetric}) 
for the symmetry level of an NCF specified using
its default-normalized representation.
We also present an efficient algorithm (Theorem~\ref{thm:rsym_canalyzing})
for the converse problem, that is,
testing whether a given $r$-symmetric function $f$ is an NCF;
if so, our algorithm constructs a default-normalized 
representation for $f$.
Next, we show that for any NCF $f$ with $n$ variables,
the two statements ``$f$ is strongly asymmetric" ~and~
``the symmetry level of $f$ is $n$"~ are equivalent 
(Theorem~\ref{thm:ncf_strong_asymmetry}).
We use this result to derive a closed form expression 
for the number of $n$-variable NCFs that are 
also strongly asymmetric (Theorem~\ref{thm:count_strongly_asymmetric}).
Our final result (Proposition~\ref{pro:ncf_symmetric})
identifies all the functions that are symmetric and
canalyzing as well as those that are symmetric and NCFs.

@@

Theorem~\ref{thm:ncf_r_symmetric} shows that given an NCF $f$,
the problem of finding the smallest integer $r$ such that
$f$ is $r$-symmetric can be solved efficiently.
We now show that the converse problem, that is, testing whether a given
$r$-symmetric Boolean function is an NCF,
can also be solved efficiently.
When the answer is ``yes", our algorithm also constructs a
default-normalized representation of $f$.

Let $f(x_1, x_2, \ldots, x_n)$ be an $r$-symmetric Boolean function of $n$
variables.
We assume that the symmetry groups $g_i$, $1 \leq i \leq r$,
of $f$ are given.
Let $m_i$ denote the number of variables
in the group containing variable $x_i$, $1 \leq i \leq m$.
Thus, in the group containing $x_i$, the number
of variables which can take on the value 1 varies from 0
to $m_i$, $1 \leq i \leq n$.
We also assume that $f$ is
given as a table $T$ with $\mu = \prod_{i=1}^r (m_i+1)$ rows.
Each row of $T$ specifies an $r$-tuple $(c_1, c_2, \ldots, c_r)$,
where $c_i$ is the number of variables of group $g_i$ which
have the value 1, along with the \{0,1\} value of $f$ for that $r$-tuple.
As will be seen, our algorithm for determining whether $f$ is
an NCF runs in $O(r^2\mu)$ time.
The algorithm relies on the following observation.

\begin{observation}\label{obs:rsym_canalyzing}
Suppose $f(x_1, x_2, \ldots, x_n)$ is an $r$-symmetric Boolean function.
Consider any variable $x_i$ and suppose
the number of variables in the group
containing $x_i$ is $m_i$, $1 \leq i \leq r$.
\begin{enumerate}
\item The function $f$ is canalyzing in variable $x_i$
with canalyzing value 1 iff all the table entries where the number
of 1's in the group containing $x_i$ is nonzero have the same value
of $f$. (This value of $f$ is the canalyzed value when $x_i = 1$.)
\item The function $f$ is canalyzing in variable $x_i$ with
canalyzing value 0 iff all the table entries where the number of
1's in the group containing $x_i$ is less than $m_i$ have the same
value of $f$. (This value of $f$ is the canalyzed value when $x_i = 0$.)
\end{enumerate}
\end{observation}
We now explain how Observation~\ref{obs:rsym_canalyzing} can be
used to develop an iterative algorithm for determining whether 
$f$ is an NCF, and if so, constructing a default-normalized 
representation (defined in Section~\ref{sec:prelim}) for $f$.
At beginning of iteration $j$, the algorithm uses a set $X_j$ of remaining variables,
table $T_j$ with $\mu_j$ rows and the number of remaining groups is $g_j$.
Initially (i.e., $j = 1$), $X_1$ = $X$ = $\{x_1, \ldots, x_n\}$,
$T_1 = T$ (the given table $T$ for $f$), 
$\mu_1 = \mu$ (the number of rows of $T$), $g_j = r$ (the number of
symmetry groups of $f$) and the NCF representation for $f$ is empty.
The algorithm carries out iteration $j$ if $g_j \geq 1$; 
in that iteration, the algorithm performs Steps I, II and III
as described below.

\smallskip
\noindent
I. Use Observation~\ref{obs:rsym_canalyzing} to determine whether there is
a canalyzing variable $x_i \in X_j$.

@@

\medskip
\noindent
The correctness of the algorithm is a direct consequence of
Observation~\ref{obs:rsym_canalyzing}. If the algorithm is successful,
it produces a simplified NCF representation of $f$; this can then be
efficiently converted into the default-normalized representation as
discussed in Section~\ref{sse:ncf_layer}.

To estimate the running time, we note that in iteration $j$, we can
determine in $O(g_j \mu_j)$ = $O(r \mu)$ time using 
Observation~\ref{obs:rsym_canalyzing}
whether there is a variable that is canalyzing.
(This can be done by considering one arbitrary variable 
from each group.)
Once a variable is identified, the other steps in that iteration
can be completed in $O(\mu_j)$ = $O(\mu)$ time.
Thus, the time for each iteration is $O(r \mu)$.
Since there are at most $r$ iterations, the running time of the
algorithm is $O(r^2 \mu)$.
The following theorem summarizes this result.

\begin{theorem}\label{thm:rsym_canalyzing}
The problem of testing whether a given 
$r$-symmetric function $f$ is an NCF can 
be solved efficiently. \QED
\end{theorem}
